# Math 365 / Comp 365: Activity 1
# Introduction to R, Polynomial Evaluation, and Programming Practice

## Charles Zhang

### Objectives
- R and R Markdown introduction
- Understand vectors, matrices, and functions in R
- Review basic programming skills (functions, for loops, if-then statements)
- Perform polynomial evaluation and test timing of various algorithms


### Part I: Vectors and Matrices in R

Here are a few different ways to create vectors in R:
```{r}
v=c(2,3,6) # this is a column vector
print(v) # just so you can see what we created
v # here is another way to see what we created
(v=c(2,3,6)) #putting parentheses around a command will also display the output
w=2:5
print(w)
y=seq(0, 10, 2) #The syntax of this is seq(from, to, by=)
print(y)
```


#### Inner and outer products

There are several different ways to "multiply" vectors $\vec{x}$ and $\vec{y}$:

1. Element-wise. Just multiply each element of $\vec{x}$ with the element in the corresponding poisiotn in $\vec{y}$. Note that $\vec{x}$ and $\vec{y}$ must have the same number of elements, and the result is a vector with the same number of elements.

2. Dot product, also known as the inner product. To compute, calculate $x \cdot y \equiv \vec{x}^T\vec{y}$ where the superscript $T$ indicates transpose. Dot product is related to the angle $\theta$ between the two vectors as $\vec{x} \cdot \vec{y} = |\vec{x}||\vec{y}| \cos \theta$. Rearranging this as $\vec{x} \cdot \vec{y}/|\vec{y}| = |\vec{x}| \cos \theta$ suggests the intuition of the dot product. It calculates a *projection* of one vector onto the other, or restated, it tells us how much of one vector is pointing in the direction of the other vector. Note that $\vec{x}$ and $\vec{y}$ must have the same number of elements, and the result is a scalar.

3. Outer product. To compute, calculate $x \otimes y \equiv \vec{x}\vec{y}^T$. Note that $\vec{x}$ and $\vec{y}$ do not need to have the same number of elements, and the result is a matrix. If $\vec{x}$ is $m \times 1$ and $\vec{y}$ is $n \times 1$ then $x \otimes y$ is $m \times n$.

4. Cross product. We are not going to worry about this one for now.

Let's calculate some examples.

```{r cache=TRUE}
x <- c(1,2,3)
y <- c(4,5,6)
z <- c(7,8,9,10)
x*y # element-wise
t(x)%*%y # dot product
sum(x*y) # dot product
x%*%t(z) # outer product
x%o%z # outer product
```



#### Creating matrices

There are also a few different ways to generate matrices in R. One is to use the commands `rbind` or `cbind` to bind together multiple row and column vectors. Here are some examples:
```{r}
(G=cbind(c(1,4,7),c(2,5,8),c(3,6,9))) # note that if you put parentheses around the entire line, the result is printed
(H=rbind(c(1,2,3),c(4,5,6),c(7,8,9)))
```

This works fine for small matrices but is not ideal if we have larger matrices with nice structured entries. Here is a way to create a matrix filled with entries of the same value:
```{r}
(Z = matrix(0, nrow = 10, ncol = 10))
```

As we saw before, can also generate matrices as the products of vectors. Here is an example:
```{r}
v1=c(-1,0,2)
v2=c(0,1,1)
M=v1%*%t(v2) #Notice the t transposes the vector. This is an outer product.
print(M)
```

What are the dimensions of the vectors in the above example? 

We'll talk about other ways to create matrices (particularly sparse matrices with many zero entries) as the semester progresses.

Once you have a matrix, you can change specific entries with square brackets after the matrix. Here are some examples:
```{r}
G[2,2]=22
print(G)
H[2,]=c(-4,-5,-6)
print(H)
```
Note that by leaving the column index blank in the second example, I'm telling R that I want to access the entire column.

To create a for loop in R, follow the syntax below:
```{r}
y=rep(0,10) #This repeats the element 0 ten times
for (i in 1:10){
  y[i]=i
}
print(y)
```
This is a silly way to create a vector, which can be done in only one line with the commands above. 

1. Now it's your turn. Generate the following matrices. You can do it with `for` loops, but try instead to generate them with just a line or two of shorter code.

a. 

$$\begin{bmatrix}
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 \\
1 & 1 & 1 & 5 & 5 & 5 & 1 \\
1 & 1 & 1 & 5 & 5 & 5 & 1 \\
1 & 1 & 1 & 1 & 1 & 1 & 1 
\end{bmatrix}$$

```{r}
A=matrix(1, nrow = 8, ncol = 7)
A[6:7,4:6]<-5
A
```

b.

$$\begin{bmatrix}
1 & 2 & 3 & \cdots & 10 \\
2 & 4 & 6 & \cdots & 20 \\
3 & 6 & 9 & \cdots & 30 \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
8 & 16 & 24 & \cdots & 80
\end{bmatrix}
$$
```{r}
(t(matrix(1:80,10,8)))
```

c.

$$\begin{bmatrix}
1 & 1 & 1 & 1 & 1 \\
4 & 4 & 4 & 4 & 4 \\
9 & 9 & 9 & 9 & 9 \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
121 & 121 & 121 & 121 & 121
\end{bmatrix}
$$

```{r}
(matrix((1:11)^2,11,5))
```

### Part II: Functions and Polynomial Evaluation


Here is an example of how to create a function in R:
```{r}
myfunction=function(x){
  y=2*x-3
  return(y)
}
```
I can evaluate the function at a specific value of $x$:
```{r}
myfunction(2)
```
or at multiple values of $x$:
```{r}
xx=3:7
myfunction(xx)
```


Now consider the polynomial $P(x)=2x^5-3x^4+x^3+10x^2-x+1$. We will evaluate $P(x)$ for $10^7$ random values three different ways.

a. First, use a for loop with standard polynomial evaluation. The for loop is written for you. You need to fill in the line that defines $P(x)$. The `system.time` command will tell you how long it takes to execute everything inside that command.

```{r}
set.seed(1234) # Set seed for random number generation
N = 10^7
x = runif(N) # Create 10^7 random values of x between 0 and 1
# You will need to code the next line and uncomment all the remaining lines
P = function(x){
  p=2*x^5-3*x^4+x^3+10*x^2-x+1
  return(p)
}
t1 <- system.time(
  for (thisx in x){
  P(thisx)
}
)[3]
 print(t1)
```

b. Now time how long it takes to evaluate the function in a vectorized manner, plugging in all values of $x$ at once rather than iterating over them in a for-loop.
```{r}
set.seed(1234) # Set seed for random number generation
P2 = function(x){
  p=2*x^5-3*x^4+x^3+10*x^2-x+1
  return(p)
}
t2 <- system.time(
  P2(x)
)[3]
 print(t2)
```
c. Now time how long it takes to evaluate the function again using a for loop, but now where the polynomial evaluation is done via Horner's method.
```{r}
set.seed(1234) # Set seed for random number generation

P3 = function(x){
  p=1+x*(-1+x*(10+x*(1+x*(-3+2*x))))
  return(p)
}
t3 <- system.time(
  for (thisx in x){
  P3(thisx)
}
)[3]
 print(t3)

```
d. Time how long it takes to evaluae the function in a vectorized manner using Horner's method for the polynomial evaluation.
```{r}
set.seed(1234) # Set seed for random number generation

P4 = function(x){
  p=1+x*(-1+x*(10+x*(1+x*(-3+2*x))))
  return(p)
}
t4 <- system.time(
  P4(x)
)[3]
 print(t4)
```
e. What is the speedup factor from the slowest case to the fastest case? (Note: the answer will vary from comptuer to computer.)

> For loop and the algorithms to calculate the polynomials.

f. Finally, implement a more general version of Horner's method. It should be of the form `Horner(coeffs,x)` and it should return the value of the polynomial whose coefficients are given by the vector `coeffs`, evaluated at the scalar (or vector of scalars) $x$. Make sure to include comments describing what your function is doing, and also test your function with sample inputs.

```{r}
set.seed(1234)
Horner<-function(coeffs,x){
  result = x[0]
  for (i in coeffs) 
        result = result*x + x[i]
    return(result)
}
t5 <- system.time(
  Horner(c(2,-3,1,10,-1,1),x)
)[3]
 print(t4)
 
```

### Part III: The Office Space Problem

If you've finished the above exercises, you can start with the following problem, which will be on HW1.

The point of this problem is twofold: (i) to illustrate what can happen if you accumulate many truncations of numbers and (ii) to give you practice writing programs.

In the 1999 movie Office Space, a character creates a program that takes fractions of cents that are truncated in a bank's transactions and deposits them into his own account. This is not a new idea, and hackers who have actually attempted it have been arrested. In this exercise, you will simulate the program to determine how long it would take to become a millionaire this way. 

Assume the following details:

* You have access to 10,000 bank accounts

* Initially, the bank accounts have values that are uniformly distributed between $100 and $100,000

* The annual interest rate on the accounts is 5%

* Interest is compounded each day and added to the accounts, except that fractions of a cent are truncated  

* The daily interest rate is thus .05/365

* The truncated fractions are deposited into an illegal account that initially has a balance of $0

* The illegal account can hold fractional values and it also accrues daily interest

Your job is to write R code that simulates this situation and finds how long it takes for the illegal account to reach a million dollars.

