# Math 365 / Comp 365: 
# Activity on Introduction to the Limits of Accuracy

### Objectives
- Practice using root-finding algorithms
- Gain exposure to issues of numerical accuracy

This activity is related to the discussion in Section 1.3.

### Example 1: *Root-finding in the presence of multiple roots*

Consider the polynomial 

$$f(x)=x^3-2x^2+\frac{4}{3}x-\frac{8}{27}=(x-\frac{2}{3})^3$$

Let's have a look at the function:
```{r}
x = seq(-1,2,len=1000)
f=function(x) {x^3-2*x^2+4/3*x-8/27}
plot(x,f(x),pch=20,main="x^3-2*x^2+4/3*x-8/27")
abline(0,0,col="red")
points(2/3,0,col="blue",pch="*")
```

a. Discuss with a neighbor to understand what each line in the plotting code is doing. If you are confused, you can use R help, see the link to graphics in R on Moodle or Google R plotting instructions.

b. Use your bisection method code to find the root. Use a starting interval of $[0,1]$ and a tolerance of $0.5*10^{-10}$. Also run this command, which will change the number of digits that R displays on output.
```{r}
options(digits=10)
bisect = function(f,interval,tol=0.5*10^-10,max.its=40){
  history = rep(NA, max.its) 
  # Insert some code here
  a<-interval[1]
  b<-interval[2]
  if(f(a)*f(b)<0) {
    while((b-a)/2>tol | max.its>0){
      c=(b+a)/2
      max.its <- max.its-1
      if(f(c)==0){
        break
      } 
      if(f(c)*f(b)<0) {a<-c} 
      if(f(c)*f(a)<0) {b<-c}
      root=c
      history=c(history,c)
    }
  } else{
    stop("f(a)f(b)<0 not satisfied")
  }
  return(list(root=root,history=history[!is.na(history)]))
}
options(digits = 10) 
interval <- c(0,1)
out=bisect(f,interval)
out
```
How close is the answer to the actual root of $\frac{2}{3}$?
```{r}
out$root
out$root-2/3
```

c. Try to find the root with the R function `uniroot`. Is that any closer?
```{r}
out2=uniroot(f,c(0,1))
out2$root
abs(out2$root-2/3)
```
d. Now let's zoom in very close to the root:
```{r}
x = seq(0.666656,0.666674,len=1000)
plot(x,f(x),pch=20,main="x^3-2*x^2+4/3*x-8/27")
abline(0,0,col="red")
points(2/3,0,col="blue",pch="*")
```

Do you have any ideas as to what might be happening? 

### Example 2: *Even when we don't have multiple roots, issues can arise, as exemplified by the Wilkinson polynomial*

The Wilkinson polynomial is named after the numerical analyst James Wilkinson.  It is defined as
$$W(x) = (x-1)(x-2)(x-3) ... (x-20),$$ 
which obviously has roots 1, 2, ..., 20. Multiplied out this becomes
$$
\begin{align}
W(x) &= 2432902008176640000 - 8752948036761600000*x + 13803759753640704000*x^2 -     12870931245150988800*x^3 \\
&+ 8037811822645051776*x^4 - 3599979517947607200*x^5  +     1206647803780373360*x^6 - 311333643161390640*x^7 \\
&+ 63030812099294896*x^8 -     10142299865511450*x^9 + 1307535010540395*x^{10} - 135585182899530*x^{11} \\
&+     11310276995381*x^{12} - 756111184500*x^{13} + 40171771630*x^{14} -1672280820*x^{15} 
+     53327946*x^{16} \\ &- 1256850*x^{17} + 20615*x^{18} - 210*x^{19} + x^{20}.\end{align}$$


Here is the function defined in two different forms and plotted
```{r}
# Function defined in factored format
Wilkinson = function(x) {
  (-20 + x)*(-19 + x)*(-18 + x)*(-17 + x)*(-16 + x)*(-15 + x)*(-14 + x)*(-13 + x)*
    (-12 + x)*(-11 + x)*(-10 + x)*(-9 + x)*(-8 + x)*(-7 + x)*(-6 + x)*(-5 + x)*(-4 + x)*
    (-3 + x)*(-2 + x)*(-1 + x)
}

# Function defined in expanded format
W = function(x){
  2432902008176640000 - 8752948036761600000*x + 13803759753640704000*x^2 - 
    12870931245150988800*x^3 + 8037811822645051776*x^4 - 3599979517947607200*x^5  + 
    1206647803780373360*x^6 - 311333643161390640*x^7 + 63030812099294896*x^8 - 
    10142299865511450*x^9 + 1307535010540395*x^10 - 135585182899530*x^11 + 
    11310276995381*x^12 - 756111184500*x^13 + 40171771630*x^14 -1672280820*x^15 + 
    53327946*x^16 - 1256850*x^17 + 20615*x^18 - 210*x^19 + x^20
}

# Plot the Wilkinson polynomial
x = seq(0,21,len=1000)
plot(x,W(x),type="l",ylim=c(-5e13,5e13),main="Wilkinson Polynomial")
abline(0,0,col="red")
```

Note that I used the `ylim` parameter to set the range that is plotted on the $y$-axis.

a. Use `bisect` and `uniroot`  to find the root at 16. Use the bracketing interval $[15.8,16.3]$ and use the unfactored form `W` of the polynomial. How many decimal places does it get correct? Note: with the factored version, it finds it exactly but that isn't so useful, because if we have a function nicely factored, we don't need to do numerical root-finding!

```{r}
interval <- c(15.8,16.3)
out=bisect(W,interval)
out
uniroot(W,interval)

```

b. This led Wilkinson to say in 1984: "Speaking for myself, I regard it as the most traumatic experience of my career as a numerical analyst." To feel his trauma, evaluate the function `W` at the root you computed in part a. What do you get? What should it be? 
```{r}
W(16.00587262)
W(16.0065135)
W(16)
```
c. Here is a new polynomial

```{r}
W2 = function(x){
  2432902008176640000 - 8752948036761600000*x + 13803759753640704000*x^2 - 
    12870931245150988800*x^3 + 8037811822645051776*x^4 - 3599979517947607200*x^5  + 
    1206647803780373360*x^6 - 311333643161390640*x^7 + 63030812099294896*x^8 - 
    10142299865511450*x^9 + 1307535010540395*x^10 - 135585182899530*x^11 + 
    11310276995381*x^12 - 756111184500*x^13 + 40171771630*x^14 -1672280820*x^15 + 
    53327946*x^16 - 1256850*x^17 + 20615*x^18 - (210+2^(-23))*x^19 + x^20
}
```
It is the same as the Wilkinson polynomial, except I have changed the coefficient of $x^{19}$ by reducing it from $-210$ to $-(210 + 2^{-23})$. This is a relatively small change in the problem. See what the root that was at $x = 20$ increases to after this change has been made.  Look for it in the interval $[19.5,20.95]$. Huge error magnification is caused by small relative errors in storing the coefficients. This is an impressive example of ill-conditioning, which we'll discuss later.

```{r}
interval <- c(19.5,20.95)
out=bisect(W2,interval)
out
uniroot(W2,interval)

```




