
## Math 365 / Comp 365: 
## Sparse Matrices Activity


Make sure to include the matrix package:
```{r,message=FALSE}
require(Matrix)
```

### Exercise 1

a) Generate the following sparse matrix, and make sure it only takes up 1560 bytes of memory.
```{r,echo=FALSE}
m6 = spMatrix(7,7, i = c(1,3:5), j = c(2,1,4,6),  x = c(pi,1.5,sqrt(2),-3))
print(m6)
object.size(m6)
```

b) Generate the following sparse matrix with just one line of R code.
```{r,echo=FALSE}
m7 = spMatrix(16,16, i = c(1:16,4:16), j = c(1:16,rep(3,13)),  x = c(rep(1,16),-13:-1))
print(m7)
```

c) Generate the following sparse matrix with just one line of R code.
```{r,echo=FALSE}
m8 = spMatrix(13,13, i = c(1:6,8:13,1:13), j = c(8:13,1:6,13:1),  x = c(11:16,6:1,seq(0,24,by=2)))
print(m8)
```

### Exercise 2

Write an `R` function `ThreeBanded(n,offset)` that generates an $n \times n$ ***sparse*** matrix with 

(i) the $i^{th}$ diagonal entry equal to $0.5+\sqrt{i}$; 

(ii) the entries on the first sub-diagonal and super-diagonal (one below and above the main diagonal) all equal to 1; 

(iii) the entries $A_{ij}$, where $|i -j|$ = offset all equal to 1; and

(iv) all other entries equal to zero. 

You can assume that $n \geq 4$ and $2 \leq$ `offset` $<n$. 

Here is an example:

`> ThreeBanded(10,5)`

$$\begin{bmatrix}
1.50 & 1.00 & & & & 1.00 &  & & & \\
1.00 & 1.91 & 1.00 & & & & 1.00  & && \\
& 1.00 & 2.23 & 1.00  & & & & 1.00  && \\
& & 1.00 & 2.50 & 1.00   & & &   & 1.00 & \\
& & &  1.00 & 2.74 & 1.00    & & && 1.00\\ 
1.00 & & & &  1.00 & 2.95 & 1.00     & &&  \\  
& 1.00  & & & & 1.00 & 3.14 & 1.00 && \\ 
&  & 1.00  & & & & 1.00 & 3.33 & 1.00 & \\   
&  &  & 1.00  & & & & 1.00 & 3.50 & 1.00 \\
&  &  & & 1.00  & & & & 1.00 & 3.66          
\end{bmatrix}$$


```{r}
ThreeBanded=function(n,offset){
  spMatrix(n, n,
           i=c(1:n, 1:(n-1), 2:n, (offset+1):n, 1:(n-offset)), 
           j=c(1:n, 2:n, 1:(n-1), 1:(n-offset), (offset+1):n),
           x=c(.5+sqrt(1:n), rep(1,(2*(2*n-1-offset)))))
}
ThreeBanded(10,5)
```




